---
title: "Species Distribution mapping examples with R"
author: "Miles McBain"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
---


```{css, echo = FALSE}
  .main-container {
    max-width: 1600px !important;
  }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source(here::here("packages.R"))
tar_load(study_species)
```


# Introduction

This is Bank St Reserve. It's a popular park in the inner north of Brisbane that
is home to a small colony of `r study_species$species` (`r study_species$vernacular_name`).
There are large signs marking the space as Koala habitat near all public entrances, but unfortunately people walking dogs
off leashes through the reserve is common. [Attacks from dogs are a major source
of Koala mortality in Greater Brisbane](https://environment.desi.qld.gov.au/wildlife/animals/living-with/koalas/threats).

```{r, fig.cap = "Banks St Reserve, Newmarket, Brisbane", fig.align = "center"}
tar_read(gg_banks_st)
```

Despite using Banks St as a weekly trail running track for the past six
years, the authors of this paper have observed a Koala but once, when it was pointed out by a
wildlife photographer. Recorded observations from the _Atlas of Living
Australia_ (ALA) for this area show why instance was potentially lucky. Koalas tend to
favour the middle of the reserve, whereas the best running is to be had around
the outside.

```{r, fig.cap = "This author's commonly used trail in pink, overlaid with hexbinned counts of recorded observations over the last 6 years", fig.align = "center"}
tar_read(gg_banks_run_and_occurrences_6yr)
```

An examination of the timstamps of the ALA sightings recorded for this location
showed they peaked in 2018 and have been in decline since. It is unclear whether
this is due to a decline in population, or a decline in interest from citizen
scientists. This question modtivated a study of the species spatial and temporal
distribution for Greater Brisbane.

```{r, fig.cap="Recorded Koala observations in Banks St reserve are in decline over recent years.", fig.align="center"}
tar_read(gg_bank_st_observation_time_series)
```

# Distribution of `r study_species$species` across Greater Brisbane

Here we present an interactive map designed to display the distribution of
`r study_species$species` across Brisbane, and look at the question of declining observarions.

The map contains many layers which can be toggled on and off using the layer control
in the top left corner. The layer initially selected displays counts of all
recorded ocurrences binned using hexagonal spatial regions. This is useful for
comparison at the suburb to city level, but detail is lost when zooming too far
either in or out.

To observe different levels of distribtion detail the reader may try viewing:

  - A point cloud layer of individual observation locations which has a 'heat
    map' like effect when zoomed out due to the application of additive colour
    blending. The points can be picked using the mouse cursor, which will
    display a tooltip containing some metadata relating to the observation.
  - A dynamically computed heat map layer, which recomputes a 2-dimensional
    kernel density when the map is zoomed. It is useful for understanding relative
    frequency of occurence given the frame of refrence defined by the currently
    zoomed map window. It displays interesting detail at all levels of zoom.

To examine the question of the temporal distribution, a series of hexagonal
binning layers were constructed each covering a 4-year span of the last 12
years. The colour scale of each layer remains relative to the highest observed hexagon
count, aiding comparison. These can be toggled individually.

```{r, fig.align="center", out.width="1080px", out.height="720px"}
tar_read(rd_brisbane_occurrences)
```

# Conclusion

Using our temporal series of hexagonal binning layers it was observed that while
observed counts of `r study_species$species` declined in the intial area of
interest at Banks St, this trend was not apparent in other 'hot spots' like the
Toohey Forest or Mount Gravatt Outlook Reserve south of the river, which showed
a marked increase in observations during the last 4-year period over the period prior.

We thus conclude that the decline of observations in Banks St Reserve is
unlikely to be associated with some general trend in declining interest from
citizen scientists. Further study of the area to assess the health of the `r study_species$species`
population is recommended.

## Reproducibility

<details><summary>Reproducibility receipt</summary>

```{r}
## datetime
Sys.time()

## repository
if(requireNamespace('git2r', quietly = TRUE)) {
  git2r::repository()
} else {
  c(
    system2("git", args = c("log", "--name-status", "-1"), stdout = TRUE),
    system2("git", args = c("remote", "-v"), stdout = TRUE)
  )
}

## session info
sessionInfo()
```

</details>
